---
title: "enmpa: Ecological Niche Modeling for Presence-absence Data"
author: "Luis F. Arias-Giraldo, Marlon E. Cobos, A. Townsend Peterson"
output: 
  github_document:
    toc: yes
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", 
                      fig.path = "man/figures/README-", out.width = "100%")
```

<!-- badges: start -->

<!-- badges: end -->

<hr>

The package `enmpa` comprises a set of tools to perform Ecological Niche
Modeling using presence-absence data. Some of the main functions help perform
data partitioning, model calibration, model selection, variable response
exploration, and model projection.

<br>

## Installation

You can install the development version of `enmpa` from
[GitHub](https://github.com/Luisagi/enmpa) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("Luisagi/enmpa")
```

<br>

## Example

### Loading packages needed

The package `terra` will be used to handle spatial data, and `enmpa` will be
used to perform ENM.

```{r message=FALSE}
library(enmpa)
library(terra)
```

<br>

### Example data

This is a basic example which shows you how to solve a common problem:

```{r message=FALSE, cache=TRUE}
# Load species occurrences and environmental data.
pa_data <- read.csv(system.file("extdata", "pa_data.csv", package = "enmpa"))
env_vars <- terra::rast(system.file("extdata", "raster_vars.tif", 
                                    package = "enmpa"))


# Presence-absence data with the values of environmental variables associated
head(pa_data)
```

Check raster layers for the projection area. Obtained from
[WorldClim](https://worldclim.org/):

-   bio_1 = Annual Mean Temperature
-   bio_12 = Annual Precipitation

```{r figures-raster_layers, fig.show="hold", out.width="70%", cache=TRUE}
terra::plot(env_vars, nr = 2)
```

<br>

### Model formulas

With `enmpa` you have the possibility to explore multiple model formulas derived
from combinations of variables considering linear (l), quadratic (q), and
product (p) responses. Product refers to pair interactions of variables.

Linear responses:

```{r cache=TRUE}
enmpa::get_formulas(dependent = "Pres_abs", 
                    independent = c("bio_1", "bio_12"), 
                    type = "l")
```

Linear + quadratic responses:

```{r cache=TRUE}
enmpa::get_formulas(dependent = "Pres_abs", 
                    independent = c("bio_1", "bio_12"), 
                    type = "lq")
```

Linear + quadratic + products responses:

```{r cache=TRUE}
enmpa::get_formulas(dependent = "Pres_abs", 
                    independent = c("bio_1", "bio_12"), 
                    type = "lqp")
```

<br>

### Model calibration and selection

The function `calibration_glm()` is a wrapper function that allows to:

-   Create model formulas
-   Fit and evaluate models based on such formulas
-   Select best performing models

Model selection consists of three steps:

1.  a first filter to keep the models with ROC AUC \>= 0.5 (statistically
    significant models).
2.  a second filter to maintain only models that meet a `selection_criterion`
    ("TSS": TSS \>= 0.4; or "ESS": maximum Accuracy - tolerance).
3.  from those, pick the ones with delta AIC \<= 2.

<br>

Results are returned as a list containing:

-   selected models `*$selected`
-   a summary of statistics for all models `*$summary`
-   results obtained from cross-validation for all models
    `*$calibration_results`

<br>

Now lets run an example of model calibration and selection:

```{r cache=TRUE}
# Linear + quadratic + products responses
cal_res <- enmpa::calibration_glm(data = pa_data,
                                  dependent = "Pres_abs",
                                  independent = c("bio_1", "bio_12"),
                                  response_type = "lpq",
                                  selection_criterion = "TSS",
                                  cv_kfolds = 5,parallel = T, n_cores = 4)
```

Process results:

```{r cache=TRUE}
# Two models were selected out of 31 models evaluated
cal_res$selected
```

<br>

### Predictions (projections)

After one or more models are selected, predictions can be made. In this case we
are projecting the model to the whole area of interest.

```{r warning=FALSE, figures-prediction_selected, fig.show="hold", out.width="70%", cache=TRUE}
# Prediction for the two selected models
preds <- enmpa::predict_selected(x = cal_res, newdata = env_vars)

# Visualization
terra::plot(preds$predictions, nr = 2)
```

<br>

### Response Curves

An important step in understanding the ecological niches that can be
characterized with these models is to explore variable responses. The following
lines of code help to do so:

```{r figures-rcurve_model_ID_1, fig.show="hold", out.width="50%", cache=TRUE}
# Response Curves for Bio_1 and Bio_2, first selected model 
par(mar = c(4, 4, .1, .1))

# BIO_1: 
enmpa::response_curve(model = preds$fitted_models$Model_ID_1,
                      variable = "bio_1",
                      new_data = env_vars)

# BIO_12:
enmpa::response_curve(model = preds$fitted_models$Model_ID_1,
                      variable = "bio_12",
                      new_data = env_vars)
```

<br>

### Variable importance

The variable importance or contribution to models can be calculated as a
function of the relative deviance explained by each predictor.

Model summary:

```{r cache=TRUE}
summary(preds$fitted_models$Model_ID_1)
```

Analysis of Deviance for the first selected model:

```{r warning=FALSE, cache=TRUE}
anova(preds$fitted_models$Model_ID_1, test = "Chi")
```

Using a function from `enmpa` you can explore variable importance in terms of
contribution.

```{r warning=FALSE, cache=TRUE}
# Relative contribution of the deviance explained
varimport <- enmpa::var_importance(preds$fitted_models$Model_ID_1)
varimport
```

Plotting these values can help with interpretations:

```{r warning=FALSE, figures-var_importance, fig.show="hold", out.width="70%", cache=TRUE}
barplot(varimport$contribution, names = varimport$predictor)
```
